---
title: "USA.gov Data from Bitly"
author: "Roberto Preste"
date: "`r Sys.Date()`"
output: github_document
editor_options: 
  chunk_output_type: inline
---

From the book (chapter 14.1):  

> In 2011, URL shortening service [Bitly](https://bitly.com/) partnered with the US government website [USA.gov](https://www.usa.gov/) to provide a feed of anonymous data gathered from users who shorten links ending with .gov or .mil. In 2011, a live feed as well as hourly snapshots were available as downloadable text files. This service is shut down at the time of this writing (2017)  

___

```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(jsonlite)
```

___

## Loading the data  

The dataset is in [JSON](https://www.json.org) format, so it is imported using the `jsonlite` package. Its `fromJSON` function however does not work in this case, since this is a NDJSON file, meaning that multiple JSON objects are contained in the same file.  
We will use the `stream_in` function to load the dataset instead.  

```{r, message=FALSE}
path <- "../datasets/bitly_usagov/example.txt"
records <- stream_in(file(path))
```

Let's convert the `records` object to a tibble.  

```{r}
records %<>% as_tibble()
```

```{r}
records
```


## Counting time zones  

We may be interested in counting the most often-occurring time zones.  

```{r}
tz_counts <- records %>% 
    count(tz) 
```

```{r}
tz_counts
```


There are some missing values, so let's first replace them.  

```{r}
records %<>% 
    mutate(tz = replace(tz, tz == "", "Missing"))
records$tz %<>% 
    replace_na("Unknown")
```

```{r}
tz_counts <- records %>% 
    count(tz) 
```

```{r}
tz_counts
```

Now we can sort entries in descending order. 

```{r}
tz_counts %<>%  
    arrange(desc(n))
```

Let's plot a bar chart with the first 10 entries from `tz_counts`.  

```{r, dpi=200}
head(tz_counts, n = 10) %>% 
    ggplot(aes(x = reorder(tz, n), y = n, fill = tz)) + 
    geom_col() + 
    coord_flip() + 
    guides(fill = F) + 
    labs(title = "Top time zones", x = "Time zone", y = "Count") 
```

Let's parse the `a` column to get some information about the most commonly used browsers.  

```{r}
records %<>%  
    mutate(browsers = str_split_fixed(records$a, " ", n = 2)[, 1])
```

```{r}
records %>% 
    count(browsers) %>% 
    arrange(desc(n))
```

Now we want to decompose time zones into Windows and non-Windows users. We can achieve that by looking at the `a` variable, checking whether it contains the `Windows` string or not. Some of the browser agents are missing, so we'll first drop rows where the `a` column is empty.  

```{r}
records_notnull <- records %>% 
    drop_na(a)
```

Then we'll create a new column, `os`, which will contain the `Win` value for Windows users and `NonWin` otherwise.  

```{r}
records_notnull %<>%  
    mutate(os = case_when(str_detect(a, "Windows") ~ "Win", 
                          TRUE ~ "NonWin"))
```

Now we'll aggregate these results to show the count of Windows and non-Windows users for each time zone.  

```{r}
agg_counts <- records_notnull %>% 
    group_by(tz, os) %>% 
    summarise(Users = n()) %>% 
    spread(os, Users, fill = 0)
```

```{r}
agg_counts
```

Let's add a `total` column that will be useful to sort our data based on the total number of users; for visualization purposes, we'll take the first 10 entries in this list.  

```{r}
agg_counts %<>%  
    mutate(total = sum(NonWin, Win)) %>% 
    arrange(desc(total))

agg_counts_top10 <- agg_counts[1:10, ]
```

In order to visualize these data, let's gather again `Win` and `NonWin` counts per each time zone.  

```{r}
count_subset <- agg_counts_top10 %>% 
    gather(key = "os", value = "users", Win, NonWin) %>% 
    arrange(tz, users) %>% 
    select(-total)
```

Now we can plot these data.  

```{r, dpi=200}
count_subset %>% 
    ggplot(aes(x = reorder(tz, users), y = users, fill = os)) + 
    geom_col(position = "dodge") + 
    coord_flip() + 
    labs(x = "Time zone", y = "Users", 
         title = "Windows and non-Windows users per time zone", fill = "OS") + 
    theme(legend.position = c(0.9, 0.14), 
          legend.background = element_rect(fill = "transparent"))
```

Let's normalize the count of users so the sum of Windows and non-Windows users is 1.  

```{r}
count_subset %<>% 
    group_by(tz, os) %>% 
    summarise(users = users) %>% 
    mutate(normed_total = users / sum(users))
```

Now we can plot these normalized data.  

```{r, dpi=200}
count_subset %>% 
    ggplot(aes(x = reorder(tz, users), y = normed_total, fill = os)) + 
    geom_col(position = "dodge") + 
    coord_flip() + 
    labs(x = "Time zone", y = "User ratio", 
         title = "Windows and non-Windows users ratio per time zone", fill = "OS") + 
    theme(legend.position = c(0.9, 0.86), 
          legend.background = element_rect(fill = "transparent"))
```

___  

```{r}
sessionInfo()
```

