---
title: "USDA Food Database"
author: "Roberto Preste"
date: "`r Sys.Date()`"
output: github_document
editor_options: 
  chunk_output_type: inline
---

From the book (chapter 14.4):  

>The US Department of Agriculture makes available a database of food nutrient information. Programmer Ashley Williams made available a version of this database in JSON format.  

___

```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(jsonlite)
```

___

## Loading the data  

This data come from the US Department of Agriculture and report a set of food nutrient information. These data were made available by Ashley Williams in JSON format, where each food has a number of identifying attributes along with two lists of nutrients and portion sizes.  

```{r, message=FALSE}
path <- "../datasets/usda_food/database.json"
db <- as_tibble(fromJSON(path, flatten = T))
```

```{r}
db
```

Some of the features in this dataset are actually nested dataframes, as in `nutrients`.  

```{r}
db[[1, "nutrients"]]
```

Without going too much in details, let's grep a subset of these information by selecting the `id`, `description`, `group` and `manufacturer` columns. In addition, we will rename the `description` and `group` columns to `food` and `fgroup`, respectively.  

```{r}
info <- db %>% 
    select(id, description, group, manufacturer) %>% 
    rename(food = description, fgroup = group)
```

Now let's see the distribution of food groups.  

```{r}
info %>% count(fgroup, sort = T)
```

Let's collect all of the nutrients into a single dataframe. First we will create a vector with the ids that will then be added to each nutrient dataframe.  

```{r}
ids <- c()
for (elem in db$id) {
    ids <- append(ids, rep(elem, dim(subset(db, id == elem)[["nutrients"]][[1]])[1]))
}
```

```{r}
nutr_rows <- bind_rows(db$nutrients)
```

```{r}
nutrients <- 
    bind_cols(id = ids, nutr_rows) %>% 
    rename(nutrient = description, nutgroup = group)
```

```{r}
nutrients
```


Now we can merge the previously created `info` dataframe with `nutrients`.  

```{r}
ndata <- full_join(info, nutrients, by = "id")
```

```{r}
ndata
```


Let's calculate median values by food group and nutrient type.  

```{r}
result <- ndata %>% 
    group_by(nutrient, fgroup) %>% 
    summarise(quant = median(value))
```

```{r}
result
```


Now we can view which foods contain most Zinc.  

```{r, dpi=200}
result %>% 
    filter(nutrient == "Zinc, Zn") %>% 
    ggplot(aes(x = reorder(fgroup, quant), y = quant, fill = fgroup)) + 
    geom_col() + 
    coord_flip() + 
    guides(fill = F) + 
    labs(x = "Food group", y = "Median value", title = "Median Zinc content")
```

___ 

```{r}
sessionInfo()
```
